Mechanized Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2023-11-16 paper-new/graduality.tex
begin rearranging: discuss term syn/sem and then ordering syn/sem

\section{Unary Canonicity}
Before discussing graduality, we seek to prove its ``unary'' analogue.
Namely, instead of considering inequality between terms, we start by considering equality.

\section{Graduality}\label{sec:graduality}
The main theorem we would like to prove is the following:

\begin{theorem}[Graduality]
  If $\cdot \vdash M \ltdyn N : \nat$, then
  \begin{enumerate}
    \item If $N = \mho$, then $M = \mho$
    \item If $N = `n$, then $M = \mho$ or $M = `n$
    \item If $M = V$, then $N = V$
  \end{enumerate}
\end{theorem}

>>>>

\section{Graduality}\label{sec:graduality}
Now we move to the proof of graduality.
The main theorem we would like to prove is the following:

% TODO use the etmprec notation
\begin{theorem}[Graduality]
  If $\cdot \vdash M \ltdyn N : \nat$, then
  \begin{enumerate}
    \item $M \Downarrow$ iff $N \Downarrow$
    \item If $M \Downarrow v_?$ and $N \Downarrow v'_?$ then either $v_? = \mho$, or $v_? = v'_?$.
  \end{enumerate}
\end{theorem}

\subsection{Proof Overview}

We split the proof into three main steps. The first concerns the translation
of the surface syntax $\extlc$ to the intensional syntax $\intlc$.

\begin{lemma}[]
  If $\etmprec{\gamlt}{M}{N}{c}$, then there exist $M'$ and $N'$ such that

  \[ i(M) \synbisim M' \ltdyn_i N' \synbisim i(N). \]
\end{lemma}

Next we show that the notions of intensional syntactic term precision and
syntactic bisimilarity imply their corresponding semantic counterparts:

\begin{lemma}[]
  If $\itmprec{\gamlt}{M}{N}{c}$, then $\sem{M} \semlt {\sem{N}}$
\end{lemma}

\begin{lemma}
  If $M \synbisim N$, then $\sem{M} \bisim \sem{N}$.
\end{lemma}

Finally, we need an adequacy result stating that the semantic notions of error
ordering and bisimilarity for natural numbers imply that the values of the
lift monad that they relate have the expected behavior.

\begin{lemma}
  Let $l, l' : \li \Nat$.
  If $l \semlt l'$, then either (1) $l = \delta^n(\mho)$ for some $n$, or (2) 
  $l = l' = \delta^n(\eta m)$ for some $n$ and $m$, or (3) $l$ and $l'$ fail to terminate.
\end{lemma}

\begin{lemma}
  If $l \bisim l'$, then either (1) $l = \delta^n(\mho)$ and $l' = \delta^{n'}(\mho)$ for some $n$ and $n'$,
  (2) $l = \delta^n(\eta m)$ and $l' = \delta^{n'}(\eta m)$, or $l$ and $l'$ fail to terminate.
\end{lemma}

### 2023-11-16 paper-new/intro.tex
begin rearranging: discuss term syn/sem and then ordering syn/sem

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
One of the principal ways of categorizing type systems of programming languages is
whether they are \emph{static} or \emph{dynamic}.
In static typing, the code is type-checked at compile time, while in dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static typing,
the programmer is assured that if the program passes the type-checker, their program
is free of type errors. Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their functions.

\emph{Gradually-typed languages} \cite{siek-taha06} allow for both static and dynamic typing disciplines
to be used in the same codebase, and support smooth interoperability between statically-typed
and dynamically-typed code.
This flexibility allows programmers to begin their projects in a dynamic style and
enjoy the benefits of dynamic typing related to rapid prototyping and easy modification
while the codebase ``solidifies''. Over time, as parts of the code become more mature
and the programmer is more certain of what the types should be, the code can be
\emph{gradually} migrated to a statically typed style without needing to start the project
over in a completely differnt language.

Gradually-typed languages should satisfy two intuitive properties.
First, the interaction between the static and dynamic components of the codebase
should be safe -- i.e., should preserve the guarantees made by the static types.
In particular, while statically-typed code can error at runtime in a gradually-typed language,
such an error can always be traced back to a dynamically-typed term that
violated the typing contract imposed by statically typed code.
% In other words, in the statically-typed portions of the codebase, type soundness must be preserved.
Second, gradually-typed langugaes should support the smooth migration from dynamic typing
to static typing, in that the programmer can initially leave off the
typing annotations and provide them later without altering the meaning of the
program.

% Graduality property
Formally speaking, gradually typed languages should satisfy the 
\emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
This property is also referred to as \emph{graduality}, by analogy with parametricity.
Intuitively, graduality says that going from a dynamic to static style should not
introduce changes in the meaning of the program.
More specifically, making the types more precise by adding typing annotations
% (replacing $\dyn$ with a more specific type
%such as integers)
will either result in the same behavior as the original, less precise program,
or will result in a type error.

\subsection{Current Approaches to Proving Graduality}
Current approaches to proving graduality include the methods of Abstracting Gradual Typing
\cite{garcia-clark-tanter2016} and the formal tools of the Gradualier \cite{cimini-siek2016}.
These allow the language developer to start with a statically typed langauge and derive a
gradually typed language that satisfies the gradual guarantee. The downside to these approaches
is that the semantics of the resulting languages are too lazy: the frameworks consider only
the $\beta$ rules and not the $\eta$ equalities. Furthermore, while these frameworks do prove
graduality of the resulting languages, they do not show the correctness of the equational theory,
which is equally important to sound gradual typing. For example, programmers often refactor their
code without thinking about whether the refactoring has broken the semantics of the program. 
It is the validity of the laws in the equational theory that guarantees that such refactorings are sound. 
Similarly, correctness of compiler optimizations rests on the validity of the corresponding equations
from the equational theory. It is therefore important that the langages that claim to be gradually typed
have provably correct equational theories.

@@

% Proving graduality via LR
New and Ahmed \cite{new-ahmed2018}
have developed a semantic approach to specifying type dynamism in terms of
\emph{embedding-projection pairs}, which allows for a particularly elegant formulation of the
gradual guarantee.
Moreover, their axiomatic account of program equivalence allows for type-based reasoning
about program equivalences.
%
In this approach, a logical relation is constructed and shown to be sound with respect to
the notion of observational approximation that specifies when one program is more precise than another.
The downside of this approach is that each new language requires a different logical relation
to prove graduality. Furthermore, the logical relations tend to be quite complicated due
to a technical requirement known as \emph{step-indexing}.
As a result, developments using this approach tend to require vast effort, with the
corresponding technical reports having 50+ pages of proofs.

An alternative approach, which we investigate in this paper, is provided by
\emph{synthetic guarded domain theory}.
The tecnhiques of synthetic guarded domain theory allow us to internalize the
step-index reasoning normally required in logical relations proofs of graduality,
ultimately allowing us to specify the logical relation in a manner that looks nearly
identical to a typical, non-step-indexed logical relation.

In this paper, we report on work we have done to mechanize proofs of graduality and
correctness of equational theories using SGDT techniques in Agda.
Our goal in this work is to mechanize these proofs in a reusable way,
thereby providing a framework to use to more easily and
conveniently prove that existing languages satsify graduality and have
sound equational theories. Moreover, the aim is for designers of new languages
to utlize the framework to facilitate the design of new provably-correct
gradually-typed languages with nontrivial features.

\subsection{Proving Graduality in SGDT}
\textbf{TODO:} This section should probably be moved to after the relevant background has been introduced?

@@

\begin{theorem}[Graduality]
  If $\cdot \vdash M \ltdyn N : \nat$, then
  \begin{enumerate}
    \item $M \Downarrow$ iff $N \Downarrow$
    \item If $M \Downarrow v_?$ and $N \Downarrow v'_?$ then either $v_? = \mho$, or $v_? = v'_?$.
  \end{enumerate}
\end{theorem}

@@

\begin{itemize}
  \item $M \ltdyn N : \nat$ means $M$ and $N$ are terms of type $\nat$ such that
  $M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
  ``more dynamic'' than $M$. Intuitively this means that $M$ and $N$ are the
  same except that in some places where $M$ has explicit typing annotations,
  $N$ has $\dyn$ instead.
  \item $\cdot \Downarrow$ is a relation on terms that is defined such that $M \Downarrow$ means
  that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.
  \item $\mho$ is a syntactic representation of a run-time type error, which
  happens, for example, when a programmer tries to call a function with a value whose type
  is found to be incompatible with the argument type of the function.
  \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
  $\mho$, or equal to the syntactic representation of a value $n$ of type $\nat$.
\end{itemize}

@@

Our first step toward proving graduality is to formulate an \emph{step-sensitive}, or \emph{intensional},
gradual lambda calculus, which we call $\intlc$, in which the computation steps taken by a term are made explicit.
The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
\emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus, denoted $\extlc$.
We will define an erasure function $\erase{\cdot} : \intlc \to \extlc$ which takes a program in the intensional lambda calculus
and ``forgets'' the syntactic information about the steps to produce a term in the extensional calculus.

Every term $M_e$ in $\extlc$ will have a corresponding program $M_i$ in $\intlc$ such that
$\erase{M_i} = M_e$. Moreover, we will show that if $M_e \ltdyn_e M_e'$ in the extensional theory,
then there exists terms $M_i$ and $M_i'$ such that $\erase{M_i}=M_e$, $\erase{M_i'}=M_e'$ and
$M_i \ltdyn_i M_i'$ in the intensional theory.

We formulate and prove an analogous graduality theorem for the intensional lambda calculus.
We define an interpretation of the intensional lambda calculus into a model in which we prove
various results. Using the observation above, given $M_e \ltdyn M_e' : \nat$, we can find
intensional programs $M_i$ and $M_i'$ that erase to them and are such that $M_i \ltdyn M_i'$.
We will then apply the intensional graduality theorem to $M_i$ and $M_i'$, and translate the result
back to $M_e$ and $M_e'$.

\subsection{Contributions}
Our main contribution is a reusable framework in Guarded Cubical Agda for developing machine-checked proofs
of graduality of a cast calculus.
To demonstrate the feasability and utility of our approach, we have used the framework to prove
graduality for the simply-typed gradual lambda calculus. Along the way, we have developed an ``intensional" theory
of graduality that is of independent interest.

In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
on synthetic guarded domain theory.
% 
In Section \ref{sec:GTLC}, we introduce the gradually-typed cast calculus
for which we will prove graduality. Important here are the notions of syntactic
type precision and term precision. We introduce both the \emph{extensional} gradual lambda calculus
($\extlc$) and the \emph{intensional} gradual lambda calculus ($\intlc$).
%
In Section \ref{sec:domain-theory}, we define several fundamental constructions
internal to SGDT that will be needed when we give a denotational semantics to
our intensional lambda calculus.
This includes the notion of Predomains as well as the concept
of EP-Pairs.
%
In Section \ref{sec:semantics}, we define the denotational semantics for the
intensional gradually-typed lambda calculus using the domain theoretic constructions in the
previous section.
%
In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
extensional gradual lambda calculus, which will make use of prove properties we prove about
 the intensional gradual lambda calculus.
%
In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
to the traditional step-indexing approach, as well as possibilities for future work.

>>>>

% gradual typing, graduality
\subsection{Gradual Typing and Graduality}
One of the principal ways of categorizing type systems of programming languages is
whether they are \emph{static} or \emph{dynamic}.
In static typing, the code is type-checked at compile time, while in dynamic typing,
the type checking is deferred to run-time. Both approaches have benefits: with static typing,
the programmer is assured that if the program passes the type-checker, their program
is free of type errors, and moreover, the equational theory implies that program optimizations are valid.
Meanwhile, dynamic typing allows the programmer to rapidly prototype
their application code without needing to commit to fixed type signatures for their functions.

\emph{Gradually-typed languages} \cite{siek-taha06, tobin-hochstadt06} allow
for both static and dynamic typing disciplines to be used in the same codebase,
and support smooth interoperability between statically-typed and dynamically-typed code.
This flexibility allows programmers to begin their projects in a dynamic style and
enjoy the benefits of dynamic typing related to rapid prototyping and easy modification
while the codebase ``solidifies''. Over time, as parts of the code become more mature
and the programmer is more certain of what the types should be, the code can be
\emph{gradually} migrated to a statically typed style without needing to
rewrite the project in a completely different language.

%Gradually-typed languages should satisfy two intuitive properties.
The following two properties have been identified as useful for gradually typed languages.
First, the interaction between the static and dynamic components of the codebase
should preserve the guarantees made by the static types.
In particular, while statically-typed code can error at runtime in a gradually-typed language,
such an error can always be traced back to a dynamically-typed term that
violated the typing contract imposed by statically typed code.
% In other words, in the statically-typed portions of the codebase, type soundness must be preserved.
Second, gradually-typed languages should support the smooth migration from dynamic typing
to static typing, in that the programmer can initially leave off the
typing annotations and provide them later without altering the meaning of the
program.

% Graduality property
Formally speaking, gradually typed languages should satisfy the 
\emph{dynamic gradual guarantee}, originally defined by Siek, Vitousek, Cimini,
and Boyland \cite{siek_et_al:LIPIcs:2015:5031}.
This property is also referred to as \emph{graduality} \cite{new-ahmed2018}, by analogy with parametricity.
Intuitively, graduality says that going from a dynamic to static style should not
introduce changes in the meaning of the program.
More specifically, making the types more precise by adding typing annotations
% (replacing $\dyn$ with a more specific type
%such as integers)
will either result in the same behavior as the original, less precise program,
or will result in a type error.

\subsection{Current Approaches to Proving Graduality}
Current approaches to proving graduality include the methods of Abstracting Gradual Typing
\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer \cite{cimini-siek2016}.
These allow the language developer to start with a statically typed language and derive a
gradually typed language that satisfies the gradual guarantee. The main downside to
these approaches lies in their inflexibility: since the process in entirely mechanical,
the language designer must adhere to the predefined framework.
Many gradually typed languages do not fit into either framework, e.g., Typed Racket
\cite{tobin-hochstadt06, tobin-hochstadt08}.
%
Furthermore, while these frameworks do prove
graduality of the resulting languages, they do not show the correctness of the equational theory,
which is equally important to sound gradual typing. For example, programmers often refactor their
code, and in so doing they rely implicitly on the validity of the laws in the equational theory.
Similarly, correctness of compiler optimizations rests on the validity of the corresponding equations
from the equational theory. It is therefore important that the languages that claim to be gradually typed
have provably correct equational theories.

% The approaches are too inflexible... the fact that the resulting semantics are too lazy
% is a consequence of that inflexibility.
% The validity of the equational theory captures the programmer's intuitive thinking when they refactor their code

@@

% Proving graduality via LR
New and Ahmed \cite{new-ahmed2018}
have developed a semantic approach to specifying type dynamism in terms of
\emph{embedding-projection pairs}, which allows for a particularly elegant formulation of the
gradual guarantee.
Moreover, their axiomatic account of program equivalence allows for type-based reasoning
about program equivalences.
%
In this approach, a logical relation is constructed and used to show that the equational theory
is sound with respect to the operational semantics.
%and shown to be sound with respect to
%the notion of observational approximation that specifies when one program is more precise than another.
The downside of this approach is that each new language requires a different logical relation
to prove graduality. Furthermore, the logical relations tend to be quite complicated due
to a technical requirement known as \emph{step-indexing}, where the stepping behavior of terms
must be accounted for in the logical relation.
As a result, developments using this approach tend to require vast effort, with the
corresponding technical reports having 50+ pages of proofs.
%
Additionally, while the axioms of gradual type theory are compositional at a ``global'' level,
they do not compose in the step-indexed setting. One of the main goals of the present work
is to formulate a composable theory of gradual typing in a setting where the stepping behavior
is tracked.

An alternative approach, which we investigate in this paper, is provided by
\emph{synthetic guarded domain theory}.
The techniques of synthetic guarded domain theory allow us to internalize the
step-index reasoning normally required in logical relations proofs of graduality,
ultimately allowing us to specify the logical relation in a manner that looks nearly
identical to a typical, non-step-indexed logical relation.

In this paper, we report on work we have done towards mechanizing proofs of graduality
and soundness of the equational theory of cast calculi using the techniques of SGDT.
We take a step toward mechanization in the Agda proof assistant by describing a compositional
denotational semantics for gradual typing in a setting where the steps taken by terms are tracked.

Our longer-term goal is to mechanize these proofs in a reusable way,
thereby providing a framework to use to more easily and
conveniently prove that existing languages satisfy graduality and have
sound equational theories. Moreover, the aim is for designers of new languages
to utilize the framework to facilitate the design of new provably-correct
gradually-typed languages with more complex features.

\subsection{Contributions}
Our main contribution is a compositional denotational semantics for step-aware gradual typing,
analogous to Gradual Type Theory but now in the ``intensional" setting.
In parallel, we are developing a reusable framework in
Guarded Cubical Agda for developing machine-checked proofs of graduality of a cast calculus.
To demonstrate the feasibility and utility of our approach, we are applying the framework
to prove graduality for the simply-typed gradual lambda calculus.
% Along the way, we have developed an ``intensional" theory of graduality that is of
% independent interest.

In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
calculus and the graduality theorem.
%
In Section \ref{sec:technical-background}, we provide technical background on gradually typed languages and
on synthetic guarded domain theory.
% 
In Section \ref{sec:gtlc-terms}, we introduce the gradually-typed cast calculus
for which we will prove graduality. We give a semantics to the terms
using the tools of guarded type theory.
% Important here are the notions of syntactic
% type precision and term precision. For reasons we describe below, we
% introduce two related calculi, one in which the stepping behavior of terms is
% implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
% is made explicit (an \emph{intensional} calculus, $\intlc$).
%

In Section \ref{sec:gtlc-precision}, we define the term precision ordering
and describe our approach to assigning a denotational semantics to this ordering.
This approach builds on the semantics constructed in the previous section,
but extending it to the term ordering presents new challenges.

% In Section \ref{sec:domain-theory}, we define several fundamental constructions
% internal to SGDT that will be needed when we give a denotational semantics to
% the gradual lambda calculus.
%This includes the Lift monad, predomains and error domains.
%
% In Section \ref{sec:semantics}, we define the denotational semantics for the
% intensional gradually-typed lambda calculus using the domain theoretic
% constructions defined in the previous section.
%
In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
extensional gradual lambda calculus.
%
In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
to the traditional step-indexing approach, as well as possibilities for future work.

\section{Overview}\label{sec:overview}

% This section used to be part of the intro.
% \subsection{Proving Graduality in SGDT}
% TODO This section should probably be moved to after the relevant background has been introduced?

@@

\begin{theorem}[Graduality]
  If $\cdot \vdash M \ltdyn N : \nat$, then
  \begin{enumerate}
    \item $M \Downarrow$ iff $N \Downarrow$.
    \item If $M \Downarrow v_?$ and $N \Downarrow v'_?$ then either $v_? = \mho$, or $v_? = v'_?$.
  \end{enumerate}
\end{theorem}

@@

\begin{itemize}
  \item $M \ltdyn N : \nat$ means $M$ and $N$ are terms of type $\nat$ such that
  $M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
  ``more dynamic'' than $M$. Intuitively this means that $M$ and $N$ are the
  same except that in some places where $M$ has explicit typing annotations,
  $N$ has $\dyn$ instead. The relation $\ltdyn$ on terms is called ``term precision''.
  Term precision is intended to model the situation where a term $M$ behaves the
  same as another term $N$ except that $M$ may error more than $N$.
  Term precision is defined by a set of axioms that capture this intuitive notion.
  This will be described in more detail in Section \ref{sec:GTLC}.

  \item $\cdot \Downarrow$ is a relation on terms that is defined such that $M \Downarrow$ means
  that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.

  \item $\mho$ is a syntactic representation of a run-time type error, which
  happens, for example, when a programmer tries to call a function with a value whose type
  is found to be incompatible with the argument type of the function.

  \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
  $\mho$, or equal to the syntactic representation of a value $n$ of type $\nat$.
\end{itemize}

@@

To prove graduality and validate the equational theory, we construct a model of the types
and terms and show that all of the axioms for term precision and for equality of terms
hold in this model. Modeling the dynamic type presents a challenge in the presence of a
language with functions: we want the dynamic type to represent a sum of all possible types
in the language, so we write down an recursive equation that the semantic object modeling
dynamic type should satisfy. When the language includes function types, this equation involves a
negative occurrence of the variable for which we are solving, and so the equation 
does not have inductive or coinductive solutions.
%
To model the dynamic type, we therefore use guarded recursion to define a suitable
semantic object that satisfies the unfolding isomorphism expected of the dynamic type.
The key is that we do not actually get an exact solution to the equation in the style
of traditional domain theory; rather, we get a ``guarded'' solution that holds ``up to a time step''.
%
That is, we introduce a notion of ``time'' and in the equation for the dynamic type,
we guard the negative occurrences of the variable by a special operator that
specifies that its argument is available ``later''.
%This can be seen as a logic that internalizes the notion of step-indexing.
See Section \ref{sec:technical-background} for more details on guarded type theory.

At a high level, the key parts of our proof are as follows:

% TODO revise this
\begin{itemize}
  \item Our first step toward proving graduality is to formulate a \emph{step-sensitive},
  or \emph{intensional}, gradual lambda calculus, which we call $\intlc$, in which the
  computation steps taken by a term are made explicit.
  The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
  \emph{surface}, \emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus,
  denoted $\extlc$.

  \item We define a translation from the surface syntax to the intensional syntax, and
  prove a theorem relating the term precision in the surface to term precision in the
  intensional syntax.

  \item We define a semantics for the intensional syntax in guarded type theory, for both the
  terms and for the term precision ordering $\ltdyn$.

\end{itemize}

### 2023-11-16 paper-new/technical-background.tex
begin rearranging: discuss term syn/sem and then ordering syn/sem

New and Licata \cite{newlicata} developed an axiomatic account of the
graduality relation on cast calculus terms and gave a denotational
model of this calculus using classical domain theory based on
$\omega$-CPOs. This semantics has scaled up to an analysis of a
dependently typed gradual calculus in \cite{asdf}. This meets our
criterion of being a reusable mathematical theory, as general semantic
theorems about gradual domains can be developed independent of any
particular syntax and then reused in many different denotational
models. However, it is widely believed that such classical domain
theoretic techniques cannot be extended to model higher-order store, a
standard feature of realistic gradually typed languages such as Typed
Racket. Thus if we want a reusable mathematical theory of gradual
typing that can scale to realistic programming languages, we need to
look elsewhere to so-called ``step-indexed'' techniques.

A series of works \cite{newahmed,newlicataahmed,newjamnerahmed}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike, classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{amalsthesis,nonpmetricparamorsomething,newjamnerahmed}. However
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions. Our goal in the current work is to extract
these reusable mathematical principles from these tedious models to
make formalization of realistic gradual languages tractible.

@@

  Because the dynamic type is modeled as a non-well-founded
  recursive type, their logical relation needs to be paramterized by natural numbers
  to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
  Reasoning about step-indexed logical relations
  can be tedious and error-prone, and there are some very subtle aspects that must
  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
  relation for the gradually-typed lambda calculus.

  In particular, the prior approach of New and Ahmed requires two separate logical
  relations for terms, one in which the steps of the left-hand term are counted,
  and another in which the steps of the right-hand term are counted.
  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
  related in both of the one-sided logical relations. Having two separate logical relations
  complicates the statement of the lemmas used to prove graduality, becasue any statement that
  involves a term stepping needs to take into account whether we are counting steps on the left
  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
  as fundamental and seemingly straightforward as transitivty.

  Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
  index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
  But this does not actually hold: we requrie that one of the two pairs of terms
  be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
  Which pair is required to satisfy this depends on which logical relation we are considering,
  (i.e., is it counting steps on the left or on the right),
  and so any argument that uses transitivity needs to consider two cases, one
  where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
  be related for all $i$. % This may not even be possible to show in some scenarios!

@@

In Ticked Cubical Type Theory \cite{TODO}, there is an additional sort
called \emph{ticks}. Given a clock $k$, a tick $t : \tick k$ serves
as evidence that one unit of time has passed according to the clock $k$.
The type $\later A$ is represented as a function from ticks of a clock $k$ to $A$.
The type $A$ is allowed to depend on $t$, in which case we write $\later^k_t A$
to emphasize the dependence.

@@

The rules for tick abstraction and application are similar to those of dependent
$\Pi$ types. 
In particular, if we have a term $M$ of type $\later^k A$, and we
have available in the context a tick $t' : \tick k$, then we can apply the tick to $M$ to get
a term $M[t'] : A[t'/t]$. We will also write tick application as $M_t$.
Conversely, if in a context $\Gamma, t : \tick k$ we have that $M$ has type $A$,
then in context $\Gamma$ we have $\lambda t.M$ has type $\later A$. % TODO dependent version?

The statements in this paper have been formalized in a variant of Agda called
Guarded Cubical Agda \cite{TODO}, an implementation of Clocked Cubical Type Theory.

>>>>

New and Licata \cite{new-licata18} developed an axiomatic account of the
graduality relation on cast calculus terms and gave a denotational
model of this calculus using classical domain theory based on
$\omega$-CPOs. This semantics has scaled up to an analysis of a
dependently typed gradual calculus in \cite{asdf}. This meets our
criterion of being a reusable mathematical theory, as general semantic
theorems about gradual domains can be developed independent of any
particular syntax and then reused in many different denotational
models. However, it is widely believed that such classical domain
theoretic techniques cannot be extended to model higher-order store, a
standard feature of realistic gradually typed languages such as Typed
Racket. Thus if we want a reusable mathematical theory of gradual
typing that can scale to realistic programming languages, we need to
look elsewhere to so-called ``step-indexed'' techniques.

A series of works \cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike, classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{amalsthesis,nonpmetricparamorsomething,new-jamner-ahmed19}. However
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions. Our goal in the current work is to extract
these reusable mathematical principles from these tedious models to
make formalization of realistic gradual languages tractible.

@@

  Because the dynamic type is modeled as a non-well-founded
  recursive type, their logical relation needs to be paramterized by natural numbers
  to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
  Intuitively, step-indexing makes the steps taken by terms observable.
  %
  Reasoning about step-indexed logical relations
  can be tedious and error-prone, and there are some very subtle aspects that must
  be taken into account in the proofs. Figure \ref{TODO} shows an example of a step-indexed logical
  relation for the gradually-typed lambda calculus.

  In particular, the prior approach of New and Ahmed requires two separate logical
  relations for terms, one in which the steps of the left-hand term are counted,
  and another in which the steps of the right-hand term are counted.
  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
  related in both of the one-sided logical relations. Having two separate logical relations
  complicates the statement of the lemmas used to prove graduality, because any statement that
  involves a term stepping needs to take into account whether we are counting steps on the left
  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
  as fundamental and seemingly straightforward as transitivity.

  Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
  index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
  But this does not actually hold: we require that one of the two pairs of terms
  be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
  Which pair is required to satisfy this depends on which logical relation we are considering,
  (i.e., is it counting steps on the left or on the right),
  and so any argument that uses transitivity needs to consider two cases, one
  where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
  be related for all $i$. % This may not even be possible to show in some scenarios!

  \begin{comment}
  As another example, the axioms that specify the behavior of casts do not hold in the
  step-indexed setting. Consider, for example, the ``lower bound'' rule for downcasting:

  \begin{mathpar}
    \inferrule*
    {\gamlt \vdash M \ltdyn N : dyn}
    {\gamlt \vdash \dn{\dyn \ra \dyn}{\dyn} M \ltdyn N}
  \end{mathpar}

  In the language of the step-indexed logical relation used in prior work, this would
  take the form

  \begin{mathpar}
    \inferrule*
    {(M, N) \in \mathcal{E}^{\sim}_{i}\sem{\dyn}}
    {(\dn{\dyn \ra \dyn}{\dyn} M, N) \in \mathcal{E}^{\sim}_{i}\sem{\injarr{}}}
  \end{mathpar}

  where $\sim$ stands for $\mathrel{\preceq}$ or $\mathrel{\succeq}$, i.e.,
  counting steps on the left or right respectively.
  %
  To show this, we would use the fact that the left-hand side steps and
  apply an anti-reduction lemma, showing that the term to which the LHS steps
  is related to the RHS where our fuel is now

  The left-hand side steps to a case inspection on $M$,
  where we unfold the recursive $\dyn$ type into a sum type and see whether the result
  is a function type.

  One way around these difficulties is to demand that the rules only hold
  ``extensionally'', i.e., we quantify universally over the step-index and
  reason about the ``global'' behavior of terms for all possible step indices.
  This is the approach taken in prior work.
\end{comment}

@@

Ticked Cubical Type Theory \cite{mogelberg-veltri2019} is an extension of Cubical
Type Theory \cite{CohenCoquandHuberMortberg2017}
in which there is an additional sort called \emph{ticks}. Given a clock $k$,
a tick $t : \tick\, k$ serves as evidence that one unit of time has passed according to the clock $k$.
The type $\later A$ is represented as a function from ticks of a clock $k$ to $A$.
The type $A$ is allowed to depend on $t$, in which case we write $\later^k_t A$
to emphasize the dependence.

@@

The rules for tick abstraction and application are similar to those of dependent
$\Pi$ types. 
In particular, if we have a term $M$ of type $\later^k A$, and we
have available in the context a tick $t' : \tick\, k$, then we can apply the tick to $M$ to get
a term $M[t'] : A[t'/t]$. We will also write tick application as $M_t$.
Conversely, if in a context $\Gamma, t : \tick\, k$ we have that $M$ has type $A$,
then in context $\Gamma$ we have $\lambda t.M$ has type $\later A$. % TODO dependent version?

The statements in this paper have been formalized in a variant of Agda called
Guarded Cubical Agda \cite{veltri-vezzosi2020}, an implementation of Clocked Cubical Type Theory.
